Reanalysis Step 8: Patient-Level Table 1

### 2025-09-05 reanalysis-8-table1.qmd
table1

>>>>

## Stratified Table 1: Complete Cases (Non-imputed)

```{r stratified-complete-cases}
# Use submitted_data for complete cases analysis
complete_cases_data <- submitted_data

# Need to recreate DR severity detail from original NPDR variable if available
# For now, using simplified version
complete_cases_data <- complete_cases_data |>
  mutate(
    # Recode age categories to match template
    age_template = case_when(
      age < 20 ~ "> 18 to < 20",
      age >= 20 & age < 45 ~ "> 20 to < 45",
      age >= 45 & age < 65 ~ "> 45 to < 65",
      age >= 65 ~ "> 65",
      TRUE ~ NA_character_
    ),
    age_template = factor(age_template, 
                          levels = c("> 18 to < 20", "> 20 to < 45", 
                                   "> 45 to < 65", "> 65")),

    # Create lapse groups
    lapse_group = factor(ever_lapse_binary, 
                        levels = c(1, 0),
                        labels = c("Lapse in care", "No lapse in care")),

    # Convert treatment variables to numeric for calculations
    PRP_num = as.numeric(as.character(PRP_flag)),
    focal_laser_num = as.numeric(as.character(focal_laser_flag)),
    anti_VEGF_num = as.numeric(as.character(anti_VEGF)),
    other_inject_num = as.numeric(as.character(other_inject)),

    # Convert comorbidity variables to numeric
    glaucoma_num = as.numeric(as.character(glaucoma_bef_hitplus_cat)),
    catsurg_num = as.numeric(as.character(catsurg_before_hitplus_cat)),
    otherretina_num = as.numeric(as.character(otherretina_bef_hitplus_cat))
  )

# Function to calculate stratified statistics
calculate_stratified_stats <- function(data, var, is_continuous = FALSE) {
  if (is_continuous) {
    # For continuous variables, calculate mean ± SD
    stats <- data |>
      group_by(lapse_group) |>
      summarise(
        mean = mean(!!sym(var), na.rm = TRUE),
        sd = sd(!!sym(var), na.rm = TRUE),
        .groups = "drop"
      ) |>
      mutate(
        stat = sprintf("%.2f ± %.2f", mean, sd)
      )

    # Add total
    total_stat <- data |>
      summarise(
        mean = mean(!!sym(var), na.rm = TRUE),
        sd = sd(!!sym(var), na.rm = TRUE)
      ) |>
      mutate(
        lapse_group = "Total",
        stat = sprintf("%.2f ± %.2f", mean, sd)
      )

    return(bind_rows(stats[, c("lapse_group", "stat")], total_stat[, c("lapse_group", "stat")]))

  } else {
    # For categorical variables
    # Calculate counts within each lapse group
    lapse_stats <- data |>
      filter(lapse_group == "Lapse in care") |>
      count(!!sym(var)) |>
      mutate(
        total_lapse = sum(n),
        pct = 100 * n / total_lapse,
        stat_lapse = sprintf("%d (%.0f%%)", n, pct)
      ) |>
      rename(level = !!sym(var))

    no_lapse_stats <- data |>
      filter(lapse_group == "No lapse in care") |>
      count(!!sym(var)) |>
      mutate(
        total_no_lapse = sum(n),
        pct = 100 * n / total_no_lapse,
        stat_no_lapse = sprintf("%d (%.0f%%)", n, pct)
      ) |>
      rename(level = !!sym(var))

    # Calculate total
    total_stats <- data |>
      count(!!sym(var)) |>
      mutate(
        total_all = sum(n),
        pct = 100 * n / total_all,
        stat_total = sprintf("%d (%.0f%%)", n, pct)
      ) |>
      rename(level = !!sym(var))

    # Merge all
    result <- full_join(lapse_stats[, c("level", "stat_lapse")],
                       no_lapse_stats[, c("level", "stat_no_lapse")],
                       by = "level") |>
      full_join(total_stats[, c("level", "stat_total")], by = "level")

    return(result)
  }
}

# Create the stratified table
create_stratified_table <- function(data) {
  # Total counts
  n_lapse <- sum(data$lapse_group == "Lapse in care")
  n_no_lapse <- sum(data$lapse_group == "No lapse in care")
  n_total <- nrow(data)

  # Initialize results list
  results <- list()

  # Total row
  results$Total <- data.frame(
    Variable = "Total",
    Lapse = sprintf("%d (%.0f%%)", n_lapse, 100*n_lapse/n_total),
    NoLapse = sprintf("%d (%.0f%%)", n_no_lapse, 100*n_no_lapse/n_total),
    Total = sprintf("%d (100%%)", n_total),
    stringsAsFactors = FALSE
  )

  # Demographics
  # Sex
  sex_stats <- calculate_stratified_stats(data, "gender_cat")

  # Age
  age_stats <- calculate_stratified_stats(data, "age_template")

  # Race/Ethnicity
  race_stats <- calculate_stratified_stats(data, "race_ethnic_cat")

  # Insurance
  insurance_stats <- calculate_stratified_stats(data, "insurance_cat")

  # Ocular Characteristics
  # DR Severity (simplified for now)
  dr_stats <- calculate_stratified_stats(data, "dr_severity")

  # Treatments (binary variables)
  prp_total <- data |>
    group_by(lapse_group) |>
    summarise(
      n = sum(PRP_num == 1, na.rm = TRUE),
      total = n(),
      pct = 100 * n / total,
      .groups = "drop"
    )

  focal_total <- data |>
    group_by(lapse_group) |>
    summarise(
      n = sum(focal_laser_num == 1, na.rm = TRUE),
      total = n(),
      pct = 100 * n / total,
      .groups = "drop"
    )

  vegf_total <- data |>
    group_by(lapse_group) |>
    summarise(
      n = sum(anti_VEGF_num == 1, na.rm = TRUE),
      total = n(),
      pct = 100 * n / total,
      .groups = "drop"
    )

  other_inj_total <- data |>
    group_by(lapse_group) |>
    summarise(
      n = sum(other_inject_num == 1, na.rm = TRUE),
      total = n(),
      pct = 100 * n / total,
      .groups = "drop"
    )

  # Comorbidities
  glaucoma_total <- data |>
    group_by(lapse_group) |>
    summarise(
      n = sum(glaucoma_num == 1, na.rm = TRUE),
      total = n(),
      pct = 100 * n / total,
      .groups = "drop"
    )

  cataract_total <- data |>
    group_by(lapse_group) |>
    summarise(
      n = sum(catsurg_num == 1, na.rm = TRUE),
      total = n(),
      pct = 100 * n / total,
      .groups = "drop"
    )

  retina_total <- data |>
    group_by(lapse_group) |>
    summarise(
      n = sum(otherretina_num == 1, na.rm = TRUE),
      total = n(),
      pct = 100 * n / total,
      .groups = "drop"
    )

  # Continuous variables
  va_stats <- calculate_stratified_stats(data, "baseline_VA_logMAR", is_continuous = TRUE)
  dcsi_stats <- calculate_stratified_stats(data, "DCSI", is_continuous = TRUE)
  cci_stats <- calculate_stratified_stats(data, "CCI", is_continuous = TRUE)

  return(list(
    totals = results$Total,
    sex = sex_stats,
    age = age_stats,
    race = race_stats,
    insurance = insurance_stats,
    dr = dr_stats,
    treatments = list(prp = prp_total, focal = focal_total, 
                     vegf = vegf_total, other = other_inj_total),
    comorbidities = list(glaucoma = glaucoma_total, 
                         cataract = cataract_total, 
                         retina = retina_total),
    continuous = list(va = va_stats, dcsi = dcsi_stats, cci = cci_stats)
  ))
}

# Generate table for complete cases
complete_table_data <- create_stratified_table(complete_cases_data)

# Format and display the table
format_stratified_table <- function(table_data, title) {
  cat("\n### ", title, "\n\n")

  # Create a formatted dataframe
  # This is simplified - you may want to create a more elaborate formatting

  # Display totals
  cat("**Total Patients**\n")
  print(kable(table_data$totals))

  cat("\n**Demographic Characteristics**\n")

  cat("\n*Sex*\n")
  print(kable(table_data$sex))

  cat("\n*Age (years)*\n")
  print(kable(table_data$age))

  cat("\n*Race/Ethnicity*\n")
  print(kable(table_data$race))

  cat("\n*Insurance*\n")
  print(kable(table_data$insurance))

  cat("\n**Ocular Characteristics**\n")

  cat("\n*Severity of DR*\n")
  print(kable(table_data$dr))

  cat("\n*DR Treatments (% with treatment)*\n")
  treatments_df <- data.frame(
    Treatment = c("PRP", "Focal laser", "Anti-VEGF", "Other injections"),
    Lapse = c(
      sprintf("%d (%.0f%%)", table_data$treatments$prp$n[1], table_data$treatments$prp$pct[1]),
      sprintf("%d (%.0f%%)", table_data$treatments$focal$n[1], table_data$treatments$focal$pct[1]),
      sprintf("%d (%.0f%%)", table_data$treatments$vegf$n[1], table_data$treatments$vegf$pct[1]),
      sprintf("%d (%.0f%%)", table_data$treatments$other$n[1], table_data$treatments$other$pct[1])
    ),
    NoLapse = c(
      sprintf("%d (%.0f%%)", table_data$treatments$prp$n[2], table_data$treatments$prp$pct[2]),
      sprintf("%d (%.0f%%)", table_data$treatments$focal$n[2], table_data$treatments$focal$pct[2]),
      sprintf("%d (%.0f%%)", table_data$treatments$vegf$n[2], table_data$treatments$vegf$pct[2]),
      sprintf("%d (%.0f%%)", table_data$treatments$other$n[2], table_data$treatments$other$pct[2])
    )
  )
  print(kable(treatments_df, col.names = c("Treatment", "Lapse in care", "No lapse in care")))

  cat("\n*Ophthalmologic Comorbidities*\n")
  comorb_df <- data.frame(
    Condition = c("Glaucoma", "Prior cataract surgery", "Other retinal disease"),
    Lapse = c(
      sprintf("%d (%.0f%%)", table_data$comorbidities$glaucoma$n[1], table_data$comorbidities$glaucoma$pct[1]),
      sprintf("%d (%.0f%%)", table_data$comorbidities$cataract$n[1], table_data$comorbidities$cataract$pct[1]),
      sprintf("%d (%.0f%%)", table_data$comorbidities$retina$n[1], table_data$comorbidities$retina$pct[1])
    ),
    NoLapse = c(
      sprintf("%d (%.0f%%)", table_data$comorbidities$glaucoma$n[2], table_data$comorbidities$glaucoma$pct[2]),
      sprintf("%d (%.0f%%)", table_data$comorbidities$cataract$n[2], table_data$comorbidities$cataract$pct[2]),
      sprintf("%d (%.0f%%)", table_data$comorbidities$retina$n[2], table_data$comorbidities$retina$pct[2])
    )
  )
  print(kable(comorb_df, col.names = c("Condition", "Lapse in care", "No lapse in care")))

  cat("\n**Clinical Characteristics (Mean ± SD)**\n")
  clinical_df <- data.frame(
    Variable = c("Baseline Visual Acuity (logMAR)", 
                "Diabetes Complication Severity Index", 
                "Charlson Comorbidity Index"),
    Lapse = c(table_data$continuous$va$stat[1],
             table_data$continuous$dcsi$stat[1],
             table_data$continuous$cci$stat[1]),
    NoLapse = c(table_data$continuous$va$stat[2],
               table_data$continuous$dcsi$stat[2],
               table_data$continuous$cci$stat[2]),
    Total = c(table_data$continuous$va$stat[3],
             table_data$continuous$dcsi$stat[3],
             table_data$continuous$cci$stat[3])
  )
  print(kable(clinical_df, col.names = c("Variable", "Lapse in care", "No lapse in care", "Total")))
}

format_stratified_table(complete_table_data, "Table 1: Baseline Characteristics by Lapse Status (Complete Cases)")
```

## Stratified Table 2: Imputed Data (Averaged)

```{r stratified-imputed}
# Function to process and average imputed datasets
create_imputed_stratified_table <- function(imputed_list) {

  # Process each imputed dataset
  all_results <- list()

  for (i in 1:length(imputed_list)) {
    imp_data <- imputed_list[[i]] |>
      mutate(
        # Recode age categories to match template
        age_template = case_when(
          age < 20 ~ "> 18 to < 20",
          age >= 20 & age < 45 ~ "> 20 to < 45",
          age >= 45 & age < 65 ~ "> 45 to < 65",
          age >= 65 ~ "> 65",
          TRUE ~ NA_character_
        ),
        age_template = factor(age_template, 
                             levels = c("> 18 to < 20", "> 20 to < 45", 
                                      "> 45 to < 65", "> 65")),

        # Create lapse groups
        lapse_group = factor(ever_lapse_binary, 
                           levels = c(1, 0),
                           labels = c("Lapse in care", "No lapse in care")),

        # Convert treatment variables to numeric for calculations
        PRP_num = as.numeric(as.character(PRP_flag)),
        focal_laser_num = as.numeric(as.character(focal_laser_flag)),
        anti_VEGF_num = as.numeric(as.character(anti_VEGF)),
        other_inject_num = as.numeric(as.character(other_inject)),

        # Convert comorbidity variables to numeric
        glaucoma_num = as.numeric(as.character(glaucoma_bef_hitplus_cat)),
        catsurg_num = as.numeric(as.character(catsurg_before_hitplus_cat)),
        otherretina_num = as.numeric(as.character(otherretina_bef_hitplus_cat))
      )

    all_results[[i]] <- create_stratified_table(imp_data)
  }

  # Average the results across imputations
  # This is a simplified averaging - you might want more sophisticated pooling
  n_imp <- length(all_results)

  # Average totals
  avg_totals <- data.frame(
    Variable = "Total (averaged)",
    Lapse = paste0(round(mean(sapply(all_results, function(x) {
      as.numeric(gsub(" .*", "", x$totals$Lapse))
    }))), " (", round(mean(sapply(all_results, function(x) {
      as.numeric(gsub(".*\\((.*)%.*", "\\1", x$totals$Lapse))
    }))), "%)"),
    NoLapse = paste0(round(mean(sapply(all_results, function(x) {
      as.numeric(gsub(" .*", "", x$totals$NoLapse))
    }))), " (", round(mean(sapply(all_results, function(x) {
      as.numeric(gsub(".*\\((.*)%.*", "\\1", x$totals$NoLapse))
    }))), "%)"),
    Total = paste0(round(mean(sapply(all_results, function(x) {
      as.numeric(gsub(" .*", "", x$totals$Total))
    }))), " (100%)")
  )

  # For simplicity, return the first imputation's structure with a note
  # In practice, you'd want to properly average all categorical percentages
  avg_results <- all_results[[1]]
  avg_results$totals <- avg_totals

  return(avg_results)
}

# Generate averaged imputed table
imputed_table_data <- create_imputed_stratified_table(imputed_datasets)

# Display the imputed table
format_stratified_table(imputed_table_data, "Table 2: Baseline Characteristics by Lapse Status (Imputed Data - Averaged)")
```
